Digging Deeper Into the State of the Practice for Domain Specific Research Software

### 2022-04-13 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Final version of digging deeper paper and conference registration and copyright transfer

	We have developed a methodology for assessing the state of the software
	development practice for a given research software domain. Our methodology
	prescribes the following steps: i) Identify the domain; ii) Identify a list
	of candidate software packages; iii) Filter the list to a length of about 30
	packages; iv) Collect repository related data on each software package, like
	number of stars, number of open issues, number of lines of code; v) Fill in
	the measurement template (the template consists of 108 questions to assess 9
	qualities (including the qualities of installability, usability and
	visibility)); vi) Rank the software using the Analytic Hierarchy Process
	(AHP); vii) Interview developers (the interview consists of 20 questions and
	takes about an hour); and, viii) Conduct a domain analysis. The collected
	data is analyzed by: i) comparing the ranking by best practices against 
	the ranking by popularity; ii) comparing artifacts, tools and processes to
	current research software development guidelines; and, iii) exploring pain
	points. We estimate the time to complete an assessment for a given domain at
	173 person hours. The method is illustrated via the example of Lattice
	Boltzmann Solvers, where we find that the top packages engaged in most of
	recommended best practices, but still show room for improvement with respect
	to providing API documentation, a roadmap, a code of conduct, programming
	style guide, uninstall instructions and continuous integration.

\keywords{research software \and software quality \and empirical measures \and analytic hierarchy process \and software artifacts \and developer pain points \and Lattice Boltzmann Method}

@@

Research software is critical for tackling problems in areas as diverse as
manufacturing, financial planning, environmental policy and medical diagnosis
and treatment.  However, developing reliable, reproducible, sustainable and fast
research software to address new problems is challenging because of the
complexity of the physical models and the nuances of floating point and parallel
computation. The importance of research software and the difficulty with its
development have prompted multiple researchers to investigate how developing
this software differs from other classes of software.  Previous studies have
focused on surveying developers
\cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}, developer interviews
\cite{Kelly2013} and case studies \cite{CarverEtAl2007,Segal2005}.  A
valuable source of information that has received less attention is the data in
publicly available software repositories.  We propose digging deeper into how
well projects are applying best practices by using a new methodology that
combines manual and automated techniques to extract repository-based
information.

The surveys used in previous studies have tended to recruit participants from
all domains of research software, possibly distinguishing them by programming
language (for instance, R developers~\cite{PintoEtAl2018}), or by the role of
the developers (for instance postdoctoral researchers \cite{UditAndKatz2017}).
Case studies, on the other hand, have focused on a few specific examples at a
time, as is the nature of case studies.  For our new methodology, we propose a
scope between these two extremes.  Rather than focus on all research software,
or just a few examples, we will focus on one domain at a time. The practical
reason for this scope is that digging deep takes time, making a broad scope
infeasible. We have imposed a practical constraint of one person month of effort
per domain.\footnote{A person month is considered to be $20$ working days ($4$
weeks in a month, with $5$ days of work per week) at $8$ person hours per day,
or $20 \cdot 8 = 160$ person hours.} Focusing on one domain at a time has more
than just practical advantages.  By restricting ourselves to a single domain we
can bring domain knowledge and domain experts into the mix.  The domain
customized insight provided by the assessment has the potential to help a
specific domain as they adopt and develop new practices.  Moreover, measuring
multiple different domains facilitates comparing and contrasting domain specific
practices.

Our methodology is built around 10 research questions.  Assuming that the domain
has been identified (Section~\ref{SecIdentifyDomain}), the first question is:

@@

We next wish to assess the representative software to determine how well they
apply current software development best practices.  At this point in the
process, to remove potential user/developer bias, we will base our assessment
only on publicly available artifacts, where artifacts are the documents, scripts
and code that we find in a project's public repository. Example artifacts
include requirements, specifications, user manuals, unit test cases, system
tests, usability tests, build scripts, API (Application Programming Interface)
documentation, READMEs, license documents, process documents, and code.
Following best practices does not guarantee popularity, so we will also compare
our ranking to how the user community itself ranks the identified projects.

@@

\begin{enumerate}
  \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do domain projects compare to research software in general with respect to the
	artifacts present in their repositories? (Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do domain projects compare to research software in general with respect to
	the use of tools (Section~\ref{Sec_CompareTools}) for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do domain projects compare to research software in general with respect to
	the processes used? (Section~\ref{Sec_CompareMethodologies})
\end{enumerate}

Only so much information can be gleaned by digging into software repos.
To gain additional insight, we need to interview developers
(Section~\ref{SecSurvey}) to learn:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on domain software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of domain developers compare to the pain points
	for research software in general? (Section~\ref{painpoints})
\end{enumerate}

Our methodology answers the research question through inspecting repositories,
using the Analytic Hierarch Process (AHP) for ranking software, interviewing
developers and interacting with at least one domain expert.  We leave the
measurement of the performance, for instance using benchmarks, to other projects
\cite{KaagstromEtAl1998}. The current methodology updates the approach used in
prior assessments of domains like Geographic Information Systems
\cite{SmithEtAl2018_arXivGIS}, Mesh Generators \cite{SmithEtAl2016}, Seismology
software \cite{SmithEtAl2018}, and statistical software for psychology
\cite{SmithEtAl2018_StatSoft}.  Initial tests of the new methodology have been
done for medical image analysis software \cite{Dong2021} and for Lattice
Boltzmann Method (LBM) software \cite{Michalski2021}.  The LBM example will be
used throughout this paper to illustrate the steps in the methodology.  The
paper ends with a summary of potential threats to validity
(Section~\ref{threats}) and our conclusions
(Section~\ref{SecConcludingRemarks}).

@@

The assessment is conducted via the following steps.  The steps depend on
interaction with a Domain Expert partner, as discussed in
Section~\ref{sec_vet_software_list}.

\begin{enumerate}
  \item Identify the domain of interest. (Section~\ref{SecIdentifyDomain})
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Collect quantitative measures from the project repositories.
	(Section~\ref{empiricalmeasures})
	\item Measure using the measurement template. (Section~\ref{empiricalmeasures})
	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
	\item Interview the developers. (Section~\ref{SecSurvey})
    \item Domain analysis. (Section~\ref{SecDomainAnalysis})
	\item Analyze the results and answer the research questions. (Sections~\ref{repmetrics}---\ref{painpoints}) %,~\ref{Sec_CompareArtifacts},~\ref{Sec_CompareTools},~\ref{Sec_CompareMethodologies},~\ref{painpoints},~\ref{Sec_AddressConcerns},
\end{enumerate}

We estimate 173 hours to complete the assessment of a given
domain~\cite{SmithEtAl2021}, which is close our goal of 160 person hours. This
estimate assumes the domain has been decided, the Domain Expert has been
recruited, and the pre-existing template spreadsheet and AHP
tool~\cite{SmithEtAl2021}, will be utilized.

@@

To be applicable for the methodology described in this document, the chosen
domain must have the following properties:

\begin{enumerate}
\item The domain must have well-defined and stable theoretical underpinning.
\item There must be a community of people studying the domain.
\item The software packages must have open source options.
\item A preliminary search, or discussion with experts, suggests that there will
  be around 30 or more candidate software packages.
\end{enumerate}

@@

The candidate software to answer~\rqref{RQ_WhatProjects} should be found through
search engine queries,  
\href{https://github.com/} {GitHub}, \href{https://swmath.org/} {swMATH}
and scholarly articles. The Domain Expert (Section~\ref{sec_vet_software_list})
should also be engaged in selecting the candidate software.  The following
properties are considered when creating the list:

@@

For the LBM example the initial list had 46 packages~\cite{Michalski2021}. This
list was filtered by scope, usage, and age to decrease the length to 24.

@@

We rank the projects by how well they follow best practices
(\rqref{RQ_HighestQuality}) via a measurement template~\cite{SmithEtAl2021}. For
each software package (each column in the template), we fill-in the rows. This
process takes about 2 hours per package, with a cap of 4 hours.  An excerpt of
the template is shown in Figure~\ref{measurement_template_image}.  In keeping
with scientific transparency, all data should be made available in a public
repository.

@@

Tools are used to find some of the measurements, such as the number of files,
number of lines of code (LOC), percentage of issues that are closed, etc. The
tool \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
each software package's GitHub repository for the number of binary files, the
number of added and deleted lines, and the number of commits over varying time
intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
(scc)} is used to measure the number of text based files as well as the number
of total, code, comment, and blank lines in each GitHub repository.

@@

Figure~\ref{Fig_OverallScore} shows the overall ranking of the LBM software
packages. In the absence of other information on priorities, the overall ranking
was calculated by assuming an equal weighting between all qualities.  The LBM data is available \href{https://data.mendeley.com/datasets/5dym63wn6z/1} {on-line}.

@@

For the current methodology time constraints necessitate a shallow domain
analysis. For each domain a table should be constructed that distinguishes the
programs under study by their variabilities.  In research software the
variabilities are often assumptions. Table~\ref{tbl_features} shows the
variabilities for LBM software~\cite{Michalski2021}.

@@

@@

To address \rqref{RQ_CompareHQ2Popular} we need to compare the ranking by best
practices to the community's ranking.  Our best practices ranking comes from the
AHP ranking (Section~\ref{empiricalmeasures}).  We estimate the community's
ranking by repository stars and watches.  The comparison will provide insight on
whether best practices are rewarded by popularity.  However, inconsistencies
between the AHP ranking and the community's ranking are inevitable for the
following reasons: 
\begin{inparaenum}[i)]
	\item the overall quality ranking via AHP makes the unrealistic assumption
	of equal weighting between quality factors;
	\item stars are known to not be a particularly good measure of popularity,
	because of how people use stars and because young projects have less time to
	accumulate stars~\cite{Szulik2017}; 
	\item and, as for consumer products, there are more factors influencing
	popularity than just quality.
\end{inparaenum}

Table~\ref{repometrics} compares the AHP ranking of the LBM package to their
popularity in the research community.  Nine packages do not use GitHub, so they
do not have a measure of stars. Looking at the stars of the other 15 packages,
we can observe a pattern where packages that have been highly ranked by our
assessment tend to have more stars than lower ranked packages. The best ranked
package by AHP (ESPResSo) has the second most stars, while the ninth ranked
package (Sailfish) has the highest number of stars. Although the AHP ranking and
the community popularity estimate are not perfect measures, they do suggest a
correlation between best practices and popularity.

\begin{table}[!ht]
	\begin{center}
		\begin{tabular}{ p{3cm}p{1.25cm}p{1.75cm}p{1.75cm}p{1.75cm}p{1.75cm} }
			\toprule
			Name & Our Ranking & Repository Stars & Repository Star Rank &
			Repository Watches & Repository Watch Rank\\
			\midrule
			ESPResSo & 1 & 145 & 2 & 19& 2\\
			Ludwig & 2 & 27 & 8 & 6& 7\\
			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
			OpenLB & 4 & No Git & No Git & No Git& No Git\\
			LUMA & 5 & 33 & 7 & 12& 4\\
			pyLBM & 6 & 95 & 3 & 10& 5\\
			DL\_MESO (LBE) & 7 & No Git & No Git & No Git & No Git\\
			Musubi & 8 & No Git & No Git & No Git & No Git\\
			Sailfish & 9 & 186 & 1 & 41& 1\\
			waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
			laboetie & 11 & 4 & 13 & 5& 8\\
			TCLB & 12 & 95 & 3 & 16& 3\\
			% MechSys & 13 & No Git & No Git & No Git& No Git\\
			% lettuce & 14 & 48 & 4 & 5& 8\\
			% ESPResSo++ & 15 & 35 & 5 & 12& 4\\
			% MP-LABS & 16 & 12 & 11 & 2& 9\\			
			% SunlightLB & 17 & No Git & No Git & No Git& No Git\\
			% LB3D & 18 & No Git & No Git & No Git& No Git\\			
			% LIMBES & 19 & No Git & No Git & No Git& No Git\\
			% LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
			% HemeLB & 21 & 12 & 11 & 12& 4\\
			% lbmpy & 22 &  11 & 12 & 2& 9\\	
			% LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
			... & ... & ... & ... & ... & ...\\				
			LatBo.jl & 24 & 17 & 10 & 8& 6\\			
			\bottomrule
		\end{tabular}
		\caption{Excerpt from Repository Ranking Metrics~\cite{Michalski2021}}
		\label{repometrics}
	\end{center}
	\end{table}

@@

We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that we observe
to those observed and recommended for research software in general.  While
filling in the measurement template (Section~\ref{empiricalmeasures}), the
domain software is examined for the presence of artifacts, which are then
categorized by frequency as: common (more than 2/3 of projects), less common
(between 1/3 and 2/3), and rare (less than 1/3). The observed frequency of
artifacts should then be compared to the artifacts recommended by research
software guidelines, as summarized in Table~\ref{Tbl_Guidelines}.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{ p{3cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm} }
\toprule
~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017}\\
\midrule
LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
README &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark\\
CONTRIBUTING &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark \\
CHANGELOG &  &  &  & \checkmark & \checkmark & & \checkmark & & \\
INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
Uninstall &  &  &  &  & & & & \checkmark & \\
Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
User manual &  &  & \checkmark &  & & & \checkmark & & \\
Tutorials &  &  &  &  & & & \checkmark & & \\
FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark\\
Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark & \\
Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark \\
Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\ 
Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark \\
Design &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark\\
API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark \\
Unit Tests & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
Test Plan &  & \checkmark &  &  & & \checkmark & & & \\
Integ. Tests &  & \checkmark & \checkmark &  & & & & \checkmark & \checkmark \\
System Tests &  & \checkmark & \checkmark &  & & \checkmark & & \checkmark & \checkmark \\
Acceptance Tests &  & \checkmark &  &  & & & & &  \\
Regression Tests &  &  & \checkmark &  & & \checkmark & & & \checkmark \\
Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark \\
Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & \\
Product Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & \\
Code of Conduct &  &  &  &  & & & \checkmark & & \\
\midrule
\end{tabular}
\caption{Commonly Recommended Artifacts in Software Development Guidelines} \label{Tbl_Guidelines}
\end{center}
\end{table}

>>>>

	To improve the productivity of research software developers we need to first
	understand their development practices.  Previous studies on this topic have
	collected data by surveying as many developers as possible, across a broad
	range of application domains.  We propose to dig deeper into the state of
	the practice by instead looking at what developers in specific domains
	create, as evidenced by the contents of their repositories.  Our methodology
	prescribes the following steps: i) Identify the domain; ii) Identify a list
	of candidate software; iii) Filter the list to a length of about 30
	packages; iv) Collect repository related data on each package, like number
	of stars, number of open issues, number of lines of code; v) Fill in the
	measurement template (the template consists of 108 questions to assess 9
	qualities (including the qualities of installability, usability and
	visibility)); vi) Rank the software using the Analytic Hierarchy Process
	(AHP); vii) Interview developers (the interview consists of 20 questions and
	takes about an hour); and, viii) Conduct a domain analysis. The collected
	data is analyzed by: i) comparing the ranking by best practices against the
	ranking by popularity; ii) comparing artifacts, tools and processes to
	current research software development guidelines; and, iii) exploring pain
	points. We estimate the time to complete an assessment at 173 person hours.
	The method is illustrated via the example of Lattice Boltzmann Solvers,
	where we find that the top packages engaged in most of recommended best
	practices, but still show room for improvement with respect to providing API
	documentation, a roadmap, a code of conduct, programming style guide and
	continuous integration.

\keywords{research software \and software quality \and empirical measures \and  software artifacts \and developer pain points \and Lattice Boltzmann Method}

@@

Research software is critical for tackling problems in areas as diverse as
manufacturing, financial planning, environmental policy and medical diagnosis
and treatment.  However, developing reliable, reproducible, sustainable and fast
research software is challenging because of the complexity of physical models
and the nuances of floating point and parallel computation. The critical
importance of research software, and the challenges with its development, have
prompted multiple researchers to investigate the state of development practice.
Understanding the creation of research software is critical for devising future
methods and tools for reducing development time and improving software quality.

Previous studies on the state of the practice for research software have often
focused on surveying
developers~\cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}. Although
surveys provide valuable information, they are limited by their reliance on what
developers say they do, rather than directly measuring what they actually do.
Therefore, we propose a state of the practice assessment methodology that
investigates the work products developers create by digging into the
software repositories they create.

Although other studies~\cite{GrannanEtAl2020,SoodEtAl2019} mine research
software repositories to estimate productivity, code quality and project
popularity, the other studies focus on automation and on code related artifacts.
To gain deeper insight, at the expense of taking more time, we will relax the
automation requirement and use manual investigation where necessary.  We will
also expand our assessment beyond code to include other artifacts, where
artifacts are the documents, scripts and code that we find in a project's public
repository. Example artifacts include requirements, specifications, user
manuals, unit test cases, system tests, usability tests, build scripts, API
(Application Programming Interface) documentation, READMEs, license documents,
and process documents.

The surveys used in previous studies have tended to recruit participants from
all domains of research software.  This is what we will call \emph{research
software in general}, as opposed to specific domain software. The surveys may
distinguish participants by programming language (for instance, R
developers~\cite{PintoEtAl2018}), or by the role of the developers (for instance
postdoctoral researchers \cite{UditAndKatz2017}), but the usual goal is to cast
as wide a net as possible. Case studies \cite{CarverEtAl2007,Segal2005}, on the
other hand, go more in depth by focusing on a few specific examples at a time.
For our new methodology, we propose a scope between these two extremes.  Rather
than focus on assessing the state of the practice for research software in
general, or just a few examples, we will focus on one scientific domain at a
time. The practical reason for this scope is that digging deep takes time,
making a broad scope infeasible. We have imposed a practical constraint of one
person month of effort per domain.\footnote{A person month is considered to be
$20$ working days ($4$ weeks in a month, with $5$ days of work per week) at $8$
person hours per day, or $20 \cdot 8 = 160$ person hours.}  Focusing on one
domain at a time has more than just practical advantages.  By restricting
ourselves to a single domain we can bring domain knowledge and domain experts
into the mix.  The domain customized insight provided by the assessment has the
potential to help a specific domain as they adopt and develop new practices.
Moreover, measuring multiple different domains facilitates comparing and
contrasting domain specific practices.

Our methodology is built around 8 research questions.  Assuming that the domain
has been identified (Section~\ref{SecIdentifyDomain}), the first question is:

@@

We next wish to assess the representative software to determine how well they
apply current software development best practices.  By \emph{best practices} we
mean methods, techniques and tools that are generally believed to improve
software development, like testing and documentation.  As we will discuss in
Section~\ref{empiricalmeasures}, we will categorize our best practices around
software qualities. Following best practices does not guarantee popularity, so
we will also compare our ranking to how the user community itself ranks the
identified projects.

@@

\begin{enumerate}
  \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do domain projects compare to research software in general with respect to the
	artifacts present in their repositories? (Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do domain projects compare to research software in general with respect to
	the use of tools? (Section~\ref{Sec_CompareTools})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do domain projects compare to research software in general with respect to
	the processes used? (Section~\ref{Sec_CompareMethodologies})
\end{enumerate}

Only so much information can be gleaned by digging into software repos. To gain
additional insight, we need to interview developers (Section~\ref{SecSurvey}),
as done in other state of the practice assessments \cite{Kelly2013}, to learn:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of domain developers compare to the pain points
	for research software in general? (Section~\ref{painpoints})
\end{enumerate}

Our methodology answers the research question through inspecting repositories,
using the Analytic Hierarch Process (AHP) for ranking software, interviewing
developers and interacting with at least one domain expert.  We leave the
measurement of the performance, for instance using benchmarks, to other projects
\cite{KaagstromEtAl1998}. The current methodology updates the approach used in
prior assessments of domains like Geographic Information Systems
\cite{SmithEtAl2018_arXivGIS}, Mesh Generators \cite{SmithEtAl2016}, Seismology
software \cite{SmithEtAl2018}, and statistical software for psychology
\cite{SmithEtAl2018_StatSoft}.  Initial tests of the new methodology have been
done for medical image analysis software \cite{Dong2021} and for Lattice
Boltzmann Method (LBM) software \cite{Michalski2021}.  The LBM example will be
used to illustrate the steps in the methodology.

@@

The assessment is conducted via the following steps, which depend on
interaction with a Domain Expert partner, as discussed in
Section~\ref{sec_vet_software_list}.

\begin{enumerate}
  \item Identify the domain of interest. (Section~\ref{SecIdentifyDomain})
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Measure using the measurement template. (Section~\ref{empiricalmeasures})
	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
	\item Interview the developers. (Section~\ref{SecSurvey})
    \item Domain analysis. (Section~\ref{SecDomainAnalysis})
	\item Analyze the results and answer the research questions.
	(Sections~\ref{SecRanking}---\ref{painpoints})
\end{enumerate}

We estimate 173 hours to complete the assessment of a given
domain~\cite{SmithEtAl2021}, which is close our goal of 160 person hours.
% This estimate assumes the domain has been decided, the Domain Expert has been
% recruited, and the pre-existing template spreadsheet and AHP
% tool~\cite{SmithEtAl2021}, will be utilized.

@@

To be applicable the chosen domain must have the following properties:

\begin{enumerate}
\item The domain must have well-defined and stable theoretical underpinnings.
\item There must be a community of people studying the domain.
\item The software packages must have open source options.
\item The domain expert says there will be at least 30 candidate packages.
\end{enumerate}

@@

The candidate software to answer \rqref{RQ_WhatProjects} should be found through
search engine queries,  
\href{https://github.com/} {GitHub}, \href{https://swmath.org/} {swMATH}
and scholarly articles. The Domain Expert (Section~\ref{sec_vet_software_list})
should also be engaged in selecting the candidate software.  The following
properties are considered when creating the list:

@@

@@

We rank the projects by how well they follow best practices
(\rqref{RQ_HighestQuality}) via a measurement template~\cite{SmithEtAl2021}. For
each software package (each column in the template), we fill-in the rows. This
process takes about 2 hours per package, with a cap of 4 hours.  An excerpt of
the template is shown in Figure~\ref{measurement_template_image}.  In keeping
with scientific transparency, all data should be made publicly available.

@@

Tools are used to find some of the measurements, such as the number of files,
number of lines of code (LOC), percentage of issues that are closed, etc. The
tool \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure each
software package's GitHub repository for the number of binary files, the number
of added and deleted lines, and the number of commits over varying time
intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
(scc)} is used to measure the number of text based files as well as the number
of total, code, comment, and blank lines in each GitHub repository.  At this
time our focus is on simple metrics, so we do no need the static analysis
capabilities of tools like
\href{https://www.hello2morrow.com/products/sonargraph} {Sonargraph} or
\href{https://www.sonarcloud.io}{SonarCloud}.

@@

@@

For each domain a table should be constructed that distinguishes the programs by
their variabilities.  In research software the variabilities are often
assumptions.

\section{Measuring and Ranking the LBM Domain} \label{SecRanking}

For the LBM example the initial list had 46 packages~\cite{Michalski2021}. To
answer \rqref{RQ_WhatProjects}, this list was filtered by scope, usage, and age
to decrease the length to 24 packages, as shown in Table~\ref{tbl_features}.
This table also shows the domain analysis, in terms of variabilities, for LBM
software~\cite{Michalski2021}.

@@

To answer \rqref{RQ_HighestQuality} the 24 LBM packages were ranked for the 9
qualities~\cite{Michalski2021}.  For space reasons we will only show the overall
ranking in Figure~\ref{Fig_OverallScore}. In the absence of information on
priorities, the overall ranking was calculated with an equal weight between all
qualities.  The LBM data is available on
\href{https://data.mendeley.com/datasets/5dym63wn6z/1} {Mendeley}.

@@

To address \rqref{RQ_CompareHQ2Popular} we need to compare the ranking by best
practices to the community's ranking.  Our best practices ranking comes from the
AHP ranking (Section~\ref{empiricalmeasures}).  We estimate the community's
ranking by repository stars and watches.  The comparison will provide insight on
whether best practices are rewarded by popularity.  However, inconsistencies
between the AHP ranking and the community's ranking are inevitable for the
following reasons: 
\begin{inparaenum}[i)]
	\item the overall quality ranking via AHP makes the unrealistic assumption
	of equal weighting between quality factors;
	\item stars are not a particularly good measure of popularity,
	because of how people use stars and because young projects have less time to
	accumulate them~\cite{Szulik2017}; 
	\item and, as for consumer products, there are more factors influencing
	popularity than just quality.
\end{inparaenum}

Table~\ref{repometrics} compares the AHP ranking of the LBM package to their
popularity in the research community.  Nine packages do not use GitHub, so they
do not have a star measure. Looking at the stars of the other 15 packages,
we can observe a pattern where packages that have been highly ranked by our
assessment tend to have more stars than lower ranked packages. The best ranked
package by AHP (ESPResSo) has the second most stars, while the ninth ranked
package (Sailfish) has the highest number of stars. Although the AHP ranking and
the community popularity estimate are not perfect measures, they do suggest a
correlation between best practices and popularity.

\begin{table}[!ht]
	\begin{center}
		\begin{tabular}{ p{3cm}p{1.25cm}p{1.75cm}p{1.75cm}p{1.75cm}p{1.75cm} }
			\toprule
			Name & Our Ranking & Repository Stars & Repository Star Rank &
			Repository Watches & Repository Watch Rank\\
			\midrule
			ESPResSo & 1 & 145 & 2 & 19& 2\\
			Ludwig & 2 & 27 & 8 & 6& 7\\
			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
			OpenLB & 4 & N/A & N/A & N/A& N/A\\
			LUMA & 5 & 33 & 7 & 12& 4\\
			pyLBM & 6 & 95 & 3 & 10& 5\\
			DL\_MESO (LBE) & 7 & N/A & N/A & N/A & N/A\\
			Musubi & 8 & N/A & N/A & N/A & N/A\\
			Sailfish & 9 & 186 & 1 & 41& 1\\
			% waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
			% laboetie & 11 & 4 & 13 & 5& 8\\
			% TCLB & 12 & 95 & 3 & 16& 3\\
			% % MechSys & 13 & N/A & N/A & N/A& N/A\\
			% lettuce & 14 & 48 & 4 & 5& 8\\
			% ESPResSo++ & 15 & 35 & 5 & 12& 4\\
			% MP-LABS & 16 & 12 & 11 & 2& 9\\			
			% SunlightLB & 17 & N/A & N/A & N/A& N/A\\
			% LB3D & 18 & N/A & N/A & N/A& N/A\\			
			% LIMBES & 19 & N/A & N/A & N/A& N/A\\
			% LB2D-Prime & 20 & N/A & N/A & N/A& N/A\\		
			% HemeLB & 21 & 12 & 11 & 12& 4\\
			% lbmpy & 22 &  11 & 12 & 2& 9\\	
			% LB3D-Prime & 23 & N/A & N/A & N/A& N/A\\	
			... & ... & ... & ... & ... & ...\\				
			LatBo.jl & 24 & 17 & 10 & 8& 6\\			
			\bottomrule
		\end{tabular}
		\caption{Excerpt from Repository Ranking Metrics~\cite{Michalski2021}}
		\label{repometrics}
	\end{center}
	\end{table}

@@

We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that we observe
to those observed and recommended for research software in general.  While
filling in the measurement template (Section~\ref{empiricalmeasures}), the
domain software is examined for the presence of artifacts, which are then
categorized by frequency as: common (more than 2/3 of projects), uncommon
(between 1/3 and 2/3), and rare (less than 1/3). The observed frequency of
artifacts should then be compared to the artifacts recommended by research
software guidelines, as summarized in Table~\ref{Tbl_Guidelines}.

As Table~\ref{Tbl_Guidelines} shows, the majority of LBM generated artifacts
correspond to general recommendations from research software developers.  Areas
where LBM developers could improve include providing: API documentation, a
product roadmap, a code of conduct, code style guide, and uninstall
instructions.
